maze-dataset

### 2025-04-08 docs/paper/paper.tex
minor edits

While maze generation itself is straightforward, the architectural
challenge comes from building a system supporting many algorithms with
configurable parameters, property filtering, and representation
transformation. This library aims to greatly streamline the process of
generating and working with datasets of mazes that can be described as
subgraphs of an \(n \times n\) lattice with boolean connections and,
optionally, start and end points that are nodes in the graph.
Furthermore, we place emphasis on a wide variety of possible text output
formats aimed at evaluating the spatial reasoning capabilities of Large
Language Models and other text-based transformer models.

@@

All implemented maze generation algorithms are stochastic by nature. For
reproducibility, the \texttt{seed} parameter of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
may be set. In practice, using provided deduplication filters, we do
find that exact duplicate mazes are generated very infrequently, even
when generating very large datasets.

@@

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes
(\protect\hyperlink{ref-ivanitskiy2023structuredworldreps}{Michael
Igorevich Ivanitskiy, Spies, et al., 2023};
\protect\hyperlink{ref-maze-dataset-arxiv-2023}{Michael Igorevich
Ivanitskiy, Shah, et al., 2023};
\protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al., 2024}).
It was extended for work on understanding the mechanisms by which
recurrent convolutional and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022}) solve mazes
given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), which required matching the pixel-padded and endpoint
constrained output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). Ongoing work using \texttt{maze-dataset} aims to
investigate the effects of varying the tokenization format on the
performance of pretrained Large Language Models (LLMs) on spatial
reasoning.

>>>>

While maze generation itself is straightforward, the architectural
challenge comes from building a system supporting many algorithms with
configurable parameters, property filtering, and representation
transformation. This library aims to greatly streamline the process of
generating and working with datasets of mazes that can be described as
subgraphs of an \(n \times n\) lattice with boolean connections and,
optionally, start and end points that are nodes in the graph.
Furthermore, we place emphasis on a wide variety of possible text output
formats aimed at evaluating the spatial reasoning capabilities of Large
Language Models (LLMs) and other text-based transformer models.

@@

All implemented maze generation algorithms are stochastic by nature. For
reproducibility, the \texttt{seed} parameter of
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
may be set. In practice, using provided deduplication filters, we find
that exact duplicate mazes are generated very infrequently, even when
generating very large datasets.

@@

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes
(\protect\hyperlink{ref-ivanitskiy2023structuredworldreps}{Michael
Igorevich Ivanitskiy, Spies, et al., 2023};
\protect\hyperlink{ref-maze-dataset-arxiv-2023}{Michael Igorevich
Ivanitskiy, Shah, et al., 2023};
\protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al., 2024}).
It was extended for work on understanding the mechanisms by which
recurrent convolutional and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022}) solve mazes
given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), which required matching the pixel-padded and endpoint
constrained output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). Ongoing work using \texttt{maze-dataset} aims to
investigate the effects of varying the tokenization format on the
performance of pretrained LLMs on spatial reasoning.
